---
title: 'Lab 03: Testing'
author: "Introduction to Statistical Computing"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
# A hook to wrap output based on a linewidth chunk option
# From https://github.com/yihui/knitr-examples/blob/master/077-wrap-output.Rmd
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE, linewidth=79)
```

Name: Yago Tobio

ICAI ID: 201802168

Collaborated with:

This lab is to be done in class (completed outside of class time). You
may collaborate with one classmate, but you must identify his/her name
above, and you must submit **your own** lab as this completed .Rmd file.

# Installing and loading packages

In order to perform the exercises in this practice you should install
and load the `testthat` and the `covr` package.

```{r}
#install.packages("covr")
#install.packages("testhat")
library(testthat)
library(covr)
```

# Q1. Unitary tests

**1a.** Let's begin with an easy one. You have the following function to
convert from Celsius to Fahrenheit. Just check that it works by using an
unitary test using the `testthat` package. You may need to see the help
for the function `?test_that`:

```{r q1a}
C_to_F <- function(C_temp){
    F_temp <- (C_temp * 9/5) + 32;
    return(F_temp);
}
# YOUR CODE GOES HERE
test_that("Celsius to Fahrenheit conversion works correctly", {
  expect_equal(C_to_F(0), 32)  # Testing the known conversion
  expect_equal(C_to_F(100), 212)  # Boiling point of water
  expect_equal(C_to_F(-40), -40)  # Point where Celsius equals Fahrenheit
})
```

**1b.** Okay, you just have learned to test a function. Review the
concept with the function to convert from Fahrenheit to Celsius:

```{r q1b}
F_to_C <- function(F_temp){
    C_temp <- (F_temp - 32) * 5/9;
    return(C_temp);
}
# YOUR CODE GOES HERE
test_that("Fahrenheit to Celsius conversion works correctly", {
  expect_equal(F_to_C(32), 0)  # Testing the known conversion
  expect_equal(F_to_C(212), 100)  # Boiling point of water
  expect_equal(F_to_C(-40), -40)  # Point where Celsius equals Fahrenheit
})
```

# Q2. Using a test file

**2a.** Test functions can be saved like any other R script file (with a
.R extension), but with one caveat. A test R script should start with
the prefix ‘test-’. A good way of doing this is to add the prefix to the
name of file that stores the functions to be tested. In this case, your
teacher have included the `C_to_F` and `F_to_C` function to the
`temp_conversion.R` file. You shall include the tests created in the
previous section to the file `test-temp_conversion.R`. After that, run
the following command to run all the test files included in the working
directory:

```{r q2a}
test_dir(".")
```

**2b.** In the test file provided there is a failing test. Correct it
and check that the `test_dir(".")` command now runs smoothly.

```{r q2b}
test_dir(".")
```

# Q3. Complicate things

**3a.** Now that you have the setup done, create a file called
`test-my_awesome_function.R` with tests for the function
`my_awesome_function()` included in `my_awesome_function.R` so that
they: - Try the function with erroneous data type, for example, using a
string instead of a number. - Try the function with vectorized inputs. -
Check that the results are numeric.

After that, modify the function in order than none of this scenarios
produce an error in the tests. You might need to include `stop()`
statements in failing cases. Try the tests using the `test_file`
function:

```{r q3a}
test_file("test-my_awesome_function.R")
```

# Q4. Code coverage

**4a.** Now that you know how to modify tests, let's obtain the code
coverage for the `number_utility.R` file. In order to do that, we will
use the `covr` package to obtain a code coverage report of this file
using the test file `test-number_utility.R`.

```{r q4a}
# SO IN ALL ESSENCE, THE COVERAGE LIBRARY WILL TELL US THROUGH THE USE OF STOP AND RETURN IN OUR FUNCTIONS. HOW MANY FEATURES WE ARE TESTING FOR. 

#install.packages(htmltools)
#install.packages(DT)
library(htmltools)
library(DT)
# Constants
number_utility <- "number_utility"
extension_r <- ".R"
extension_htm <- ".htm"

# Manage file names
code_file_name <- paste(number_utility, extension_r, sep="") #number_utility.R
test_file_name <- paste("test-", code_file_name, sep="")   #test-number_utility.R
coverage_report_file_name <- paste("coverage_report_", number_utility, extension_htm, sep="")
#^^This generates the HTML file, with the coverage report for our testing. 

# Run tests and generate Code Coverage Report
test_file(test_file_name)
res <- file_coverage(code_file_name, test_file_name)
print(res)
report(res, coverage_report_file_name)
# Got it to 100% Coverage
```

The code coverage is a sad 47.37% using the current tests. Modify the
test file in order to achieve at least a 60% of code coverage.
