---
title: 'Lab 01: Effective Programming'
author: "Introduction to Statistical Computing"
output:
  pdf_document: default
  html_document: default
---

```{r, include=FALSE}
# A hook to wrap output based on a linewidth chunk option
# From https://github.com/yihui/knitr-examples/blob/master/077-wrap-output.Rmd
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE, linewidth=79)
```

Name: Victor Ovejero ICAI ID: 201904136 Collaborated with: Yago Tobio, Luis Bueno

This lab is to be done in class (completed outside of class time). You may collaborate with one classmate, but you must identify his/her name above, and you must submit **your own** lab as this completed .Rmd file.

# Installing and loading packages

In order to perform the exercise in this practice you should install and load the `microbenchmark` and `profvis` packages. Also install `devtools` in order to install the `efficientR` package where there are datasets that are used in this practice using `devtools::install_github("csgillespie/efficientR")`. You should also install the `proftools` package from CRAN.

```{r}
# YOUR CODE GOES HERE
```

From the `Bioconductor` repository you must also install the `graph` and `Rgraphviz` packages. To install packages from this repository, you must install `BiocManager` package first and then use the `BiocManager::install()` function to install the packages.

```{r, eval = FALSE}
install.packages("BiocManager", dep = TRUE)
BiocManager::install(c("Rgraphviz","graph"))
```

# Q1. Microbenchmarking

**1a.** Use the `microbenchmark::microbenchmark()` function to know which of the following three functions is the fastest to perform the cumulative sum of a 100-element vector. By how much is the fastest with respect to the second one?

```{r q1a, eval=FALSE}
x <- 1:100 # initiate vector to cumulatively sum

# Method 1: with a for loop (10 lines)
cs_for <- function(x) {
  for (i in x) {
    if (i == 1) {
      xc = x[i]
    } else {
      xc = c(xc, sum(x[1:i]))
    }
  }
  xc
}

# Method 2: with apply (3 lines)
cs_apply <- function(x) {
  sapply(x, function(x) sum(1:x))
}

# Method 3: cumsum (1 line)
cumsum(x)

# YOUR CODE GOES HERE

library(microbenchmark)

#Microbenchmark solo permite la comparacion de 2 funciones a la vez. Con un default de 100 evaluaciones hechas en ns.

microbenchmark(cs_for(x),cs_apply(x))
microbenchmark(cs_for(x),cumsum(x))
microbenchmark(cs_apply(x),cumsum(x))
```

**1b.** Run the same benchmark but now `x` is `1:50000`. As the benchmark could take too long, set the argument `time = 1` in the `microbenchmark` function. Does the relative difference between the fastest and the second fastest increase or decrease? By how much?

```{r q1b}
# YOUR CODE GOES HERE
```

**1c.** Try profiling a section of code you have written using the `profvis::profvis()` function. Where are the bottlenecks? Were they where you expected?

```{r q1c}
# YOUR CODE GOES HERE
```

**1d.** Let's profile a section of code with the `Rprof()` function. The code section is a function to compute sample variance of a numeric vector:

```{r}
# Compute sample variance of numeric vector x
sampvar <- function(x) {
    # Compute sum of vector x
    my.sum <- function(x) {
        sum <- 0
        for (i in x) {
            sum <- sum + i
        }
        sum
    }
   
    # Compute sum of squared variances of the elements of x from
    # the mean mu
    sq.var <- function(x, mu) {
        sum <- 0
        for (i in x) {
            sum <- sum + (i - mu) ^ 2
        }
        sum
    }
   
    mu <- my.sum(x) / length(x)
    sq <- sq.var(x, mu)
    sq / (length(x) - 1)
}
```

To use the `Rprof()` function, you shall specify in which file you want to store the results of the profiling. Then you execute the code you want to profile, and then you execute `Rprof(NULL)` to stop profiling. In order to profile the `sampvar()` function applied to a random 100 million number vector:

```{r, eval = FALSE}
x <- runif(1e8)
Rprof("Rprof.out", memory.profiling = TRUE)
y <- sampvar(x)
Rprof(NULL)
```

Use the `summaryRprof()` function to print a summary of the code profiling. Which part of the function takes more time to execute? Which part of the function requires more memory?

```{r q1d}
# YOUR CODE GOES HERE
```

**1e.** `summaryRprof()` function prints a summary of the code profiling, but it is not user-friendly to read. Using the `proftool` packages, let's see the results from the `Rprof.out` file. See the help (`?`) for the functions `readProfiledata` and `plotProfileCallGraph` and plot the results of the code profiling from **1d**.

```{r q1e}
# YOUR CODE GOES HERE
```

# Q2. Efficient set-up

Let's check if you have an optimal R installation.

**2a.** What is the exact version of your computer's operating system?

```{r q2a}
# YOUR CODE GOES HERE
```

**2b.** Start an activity monitor and execute the following chunk.In it, `lapply()` (or its parallel version `mclapply()`) is used to apply a function, median(), over every column in the data frame object X

```{r q2b, eval=FALSE}
# Note: uses 2+ GB RAM and several seconds or more depending on hardware
# 1: Create large dataset
X <- as.data.frame(matrix(rnorm(1e9), nrow = 1e8))
# 2: Find the median of each column using a single core
r1 <- lapply(X, median)
# 3: Find the median of each column using many cores
r2 <- parallel::mclapply(X, median)
```

**2c.** Try modifying the settings of your RStudio setup using the Tools \> Global Options menu. What settings do you think can affect R performance? (Note only some of them, not ALL of them)

```{r q2c}
# YOUR CODE GOES HERE
```

**2d.** Try some of the shortcuts integrated in RStudio. What shortcuts do you think can save you development time? (Note only some of them, not ALL of them)

```{r q2d}
# YOUR CODE GOES HERE
```

**2e.** Check how well your computer is suited to perform data analysis tasks. In the following code chunk you will run a benchmark test from the `benchmarkme` package and plot your result against the results from people around the world. Do you think that you should upgrade your computer?

```{r q2e, eval=FALSE }
library("benchmarkme")
# Run standard tests
res_std <- benchmark_std(runs=3)
plot(res_std)
# Run memory I/O tests by reading/writing a 5MB file
res_io <- benchmark_io(runs = 1, size = 5)
plot(res_io)
```

# Q3. Efficient programming

**3a.** Create a vector `x` of 100 random numbers and use the microbenchmark package to compare the vectorised construct `x = x + 1` to the for loop version `for (i in seq_len(n)) x[i] = x[i] + 1`. Try varying the size of the input vector and check how the results differ. Which functions are being called by each method?

```{r q3a}
# YOUR CODE GOES HERE
```

**3b.** Monte Carlo integration can be performed with the following code:

```{r, eval=FALSE}
monte_carlo = function(N) {
  hits = 0
  for (i in seq_len(N)) {
    u1 = runif(1)
    u2 = runif(1)
    if (u1 ^ 2 > u2)
      hits = hits + 1
  }
  return(hits / N)
}
```

Create a vectorized function `monte_carlo_vec` which do not use a `for` loop.

```{r q3b}
# YOUR CODE GOES HERE
```

**3c.** How much faster is the vectorized function `monte_carlo_vec` with respect to the original function `monte_carlo`?

```{r q3c}
# YOUR CODE GOES HERE
```

**3d.** Using the `memoise` function, create a function called `m_fib` that is the memoized version of the recursive function:

```{r, eval=FALSE}
fib <- function(n)  {
  if(n == 1 || n == 2) return(1)
  fib(n-1) + fib(n-2)
}
```

Then, using `microbenchmark`, simulate calculating the 10th position of the Fibonacci serie a 100 times with each function. How much faster is the memoized version?

```{r q3d}
# YOUR CODE GOES HERE
```

**3e.** Try varying the parameters of the **3d** exercise. What happens when you measure the computing time of calculating the 1st position of Fibonacci serie? And the 25th?

```{r q3e}
# YOUR CODE GOES HERE
```

**3f.** Create the `c_fib` function as the compilation version of the `fib` function declared in exercise **3d** using the `cmpfun` of the `compiler` package. Which is faster, `fib`, `c_fib` or `m_fib`? And `cm_fib` (compiled version of `m_fib`)? And `mc_fib` (memoized version of `c_fib`)?

```{r q3f}
# YOUR CODE GOES HERE
```

**Challenge 01.** Calculate the computing time for calculating the Fibonacci serie 5 times from the 1st to the 25th position with the `fib`, `c_fib`, `m_fib`, `cm_fib` and `mc_fib` functions. Store the results for each position and create a plot showing these results. When does it begin to compensate using the memoized version? Hint: Use `geom_point()` and `geom_errorbars()` function of `ggplot2` to show the `median`, `lq` and `uq` values of the `microbenchmark` analysis.

```{r q3ch1}
# YOUR CODE GOES HERE
```

# Q4. Efficient data I/O

**4a.** Import data from <https://github.com/mledoze/countries/raw/master/countries.json> using the `import()` function from the `rio` package.

```{r q4a}
# YOUR CODE GOES HERE
```

**4b.** Export the data imported in **4a** to 3 different file formats of your choosing supported by `rio` (see `vignette("rio")` for supported formats). Try opening these files in external programs. Which file formats are more portable?

```{r q4b}
# YOUR CODE GOES HERE
```

**Challenge 03.** Create a simple benchmark to compare the write times for the different file formats of **4b**. Which is fastest? Which is the most space efficient?

```{r q4ch3}
# YOUR CODE GOES HERE
```

# Q5. Efficient data carpentry

**5a.** Create the following data.frame:

```{r, eval = FALSE}
df_base = data.frame(colA = "A")
```

Try and guess the output of the following commands. Quit the `eval = FALSE` argument and check if the output is what you thought.

```{r, eval = FALSE}
print(df_base)
df_base$colA
df_base$col
df_base$colB
```

Now create a tibble tbl_df and repeat the above commands.

```{r q5a}
# YOUR CODE GOES HERE
```

**5b.** Load and look at subsets of the `pew` and `lnd_geo_df` datasets from the `efficient` package. What is untidy about them? Convert each of these datasets into tidy form.

```{r q5b}
# YOUR CODE GOES HERE
```

**5c.** Consider the following string of phone numbers and fruits:

```{r, eval = FALSE}
strings = c(" 219 733 8965", "329-293-8753 ", "banana", "595 794 7569",
            "387 287 6718", "apple", "233.398.9187  ", "482 952 3315", 
            "239 923 8115", "842 566 4692", "Work: 579-499-7527", 
            "$1000", "Home: 543.355.3679")
```

Write expressions in `stringr` and `base R` that return a logical vector reporting whether or not each string contains a number.

```{r q5c}
# YOUR CODE GOES HERE
```

# Q6. Efficient optimization

**6a.** Create a vector `x` and benchmark `any(is.na(x))` against `anyNA(x)`. Do the results vary with the size of the vector?

```{r q6a}
# YOUR CODE GOES HERE
```

**6b.** Construct a `matrix` of `integer`s and a `matrix` of `numeric`s and use a `pryr::object_size()` to compare the object occupation.

```{r q6b}
# YOUR CODE GOES HERE
```

**6c.** Consider the following piece of code:

```{cpp, eval = FALSE}
double test1() {
  double a = 1.0 / 81;
  double b = 0;
  for (int i = 0; i < 729; ++i)
    b = b + a;
  return b;
}
```

-   Save the function `test1()` in a separate file. Make sure it works.
-   Write a similar function in R and compare the speed of the C++ and R versions.
-   Create a function called `test2()` where the `double` variables have been replaced by `float`. Do you still get the correct answer?
-   Change `b = b + a` to `b += a` to make your code more `C++` like.
-   (Bonus) What's the difference between `i++` and `++i`?

# Q7. Efficient hardware

**7a.** How much RAM does your computer have? (Optional question, privacy above all. Write a random number if you do not want to share your hardware information.)

```{r q7a}
# YOUR CODE GOES HERE
```

**7b.** Using your preferred search engine, how much does it cost to double the amount of available RAM on your system? (Again, write a random number if you do not want to share your hardware information)

```{r q7b}
# YOUR CODE GOES HERE
```

**7c.** Check if you are using a 32-bit or 64-bit version of R.

```{r q7c}
# YOUR CODE GOES HERE
```
