---
title: 'Lab 0.1: dplyr'
author: "Introduction to Statistical Computing"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
# A hook to wrap output based on a linewidth chunk option
# From https://github.com/yihui/knitr-examples/blob/master/077-wrap-output.Rmd
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE, linewidth=79)
```

Name: **Equipo 1**

ICAI ID: **201802168, 201810661, 201904136, 201910180**

Collaborated with: **Yago Tobio, Luis Bueno, Victor Ovejero, Jesús
López**

This lab is to be done in class (completed outside of class time). You
may collaborate with one classmate, but you must identify his/her name
above, and you must submit **your own** lab as this completed .Rmd file.

# Installing and loading packages

Below we install `tidyverse` which gives us the packages we need
(`purrr` and `dplyr`) needed to complete this lab. We also install the
`repurrrsive` package which has the Game of Thrones data set that we'll
use for the first couple of questions. Since this may be the first time
installing packages for some of you, we'll show you how. If you already
have these packages installed, then you can of course skip this part.
Note: *do not remove `eval=FALSE` from the above code chunk*, just run
the lines below in your console. You can also select "Tools" --\>
"Install Packages" from the RStudio menu.

```{r, eval=FALSE}
install.packages("tidyverse")
install.packages("repurrrsive")
```

Now we'll load the packages we need. Note: the code chunk below will
cause errors if you try to knit this file without installing the
packages first.

```{r, warning = FALSE}
library(purrr)
library(dplyr)
library(tidyr)
library(repurrrsive)
```

# Q1. Pipes to base R

For each of the following code blocks, which are written with pipes,
write equivalent code in base R (to do the same thing).

**1a.**

```{r}
letters %>%
  toupper %>%
  paste(collapse="+") 
```

```{r q1a}
paste(toupper(letters), collapse = '+')
```

**1b.**

```{r}
"     Ceci n'est pas une pipe     " %>% 
  gsub("une", "un", .) %>%
  trimws
```

```{r q1b}
trimws(gsub("une", "un", "     Ceci n'est pas une pipe     "))
```

**1c.**

```{r}
rnorm(1000) %>% 
  hist(breaks=30, main="N(0,1) draws", col="pink", prob=TRUE) 
```

```{r q1c}
hist(rnorm(1000), breaks = 30, main = "N(0,1) draws", col = "pink", prob = TRUE)
```

**1d.**

```{r}
rnorm(1000) %>% 
  hist(breaks=30, plot=FALSE) %>%
  `[[`("density") %>%
  max
```

```{r q1d}
max(hist(rnorm(1000), breaks = 30, plot = FALSE)$density)
```

# Q2. Base R to pipes

For each of the following code blocks, which are written in base R,
write equivalent code with pipes (to do the same thing).

**2a.** Hint: you'll have to use the dot `.`, as seen above in Q1b, or
in the lecture notes.

```{r}
paste("Your grade is", sample(c("A","B","C","D","R"), size=1))
```

```{r q2a}
sample(c("A", "B", "C", "D", "R"), size = 1) %>%
  paste("Your grade is", .)
```

**2b.** Hint: you can use the dot `.` again, in order to index
`state.name` directly in the last pipe command.

```{r}
state.name[which.max(state.x77[,"Illiteracy"])] 
```

```{r 2qb}
state_illiteracy <- tibble(StateName = state.name, Illiteracy = state.x77[,"Illiteracy"]) %>%
  arrange(-Illiteracy) %>%
  head(1) %>%
  pull(StateName)

print(state_illiteracy)
```

**2c.** Note: `str.url` is defined for use in this and the next
question; you can simply refer to it in your solution code (it is not
part of the code you have to convert to pipes).

```{r}
str.url = "http://www.stat.cmu.edu/~ryantibs/statcomp/data/king.txt"

lines = readLines(str.url)
text = paste(lines, collapse=" ")
words = strsplit(text, split="[[:space:]]|[[:punct:]]")[[1]]
wordtab = table(words)
wordtab = sort(wordtab, decreasing=TRUE)
head(wordtab, 10)
```

```{r q2c}
str.url <- "http://www.stat.cmu.edu/~ryantibs/statcomp/data/king.txt"

wordtab <- str.url %>%
  readLines() %>%
  paste(collapse = " ") %>%
  strsplit(split = "[[:space:]]|[[:punct:]]") %>%
  `[[`(1) %>%
  table() %>%
  sort(decreasing = TRUE)

head(wordtab, 10)
```

**2d.** Hint: the only difference between this and the last part is the
line `words = words[words != ""]`. This is a bit tricky line to do with
pipes: use the dot `.`, once more, and manipulate it as if were a
variable name.

```{r}
lines = readLines(str.url)
text = paste(lines, collapse=" ")
words = strsplit(text, split="[[:space:]]|[[:punct:]]")[[1]]
words = words[words != ""]
wordtab = table(words)
wordtab = sort(wordtab, decreasing=TRUE)
head(wordtab, 10)
```

```{r q2d}
wordtab <- str.url %>%
  readLines() %>%
  paste(collapse = " ") %>%
  strsplit(split = "[[:space:]]|[[:punct:]]") %>%
  pluck(1) %>%
  discard(~ . == "") %>%
  table() %>%
  sort(decreasing = TRUE)

head(wordtab, 10)
```

# Q3. Warming up with map

**3a.** Using the map functions from the `purrr` package, extract the
names of the characters in `got_chars` so that you produce a character
vector of length 30. Do this four different ways: (i) using `map()`,
defining a custom function on-the-fly, and casting the resulting list
into an appropriate data structure; (ii) using one of the `map_***()`
functions, but still defining a custom function on-the-fly; (iii) using
one of the `map_***()` functions, and using one of `` `[`() `` or
`` `[[`() `` functions, as well as an additional argument; (iv) using
one of the `map_***()` functions, and passing a string instead of a
function (relying on its ability to define an appropriate extractor
accordingly).

Store each of the results in a different vector and check that they are
all identical.

```{r q3a}
library(purrr)

names1 <- map(got_chars, ~ .x$name) %>% unlist()

names2 <- map_chr(got_chars, ~ .x$name)

names3 <- map_chr(got_chars, `[[`, "name")

names4 <- map_chr(got_chars, "name")

identical(names1, names2) # Should return TRUE
identical(names2, names3) # Should return TRUE
identical(names3, names4) # Should return TRUE
```

**3b.** Produce an integer vector that represents how many allegiances
each character holds. Do this with whichever map function you'd like,
and print the result to the console. Then use this (and your a saved
object from the last question) to answer: which character holds the most
allegiances? The least?

```{r q3b}
# Create an integer vector of allegiances count
allegiances_count <- map_int(got_chars, function(x) length(x$allegiances))

# Print the result
print(allegiances_count)

# Extract the names and allegiances count into a data frame
<<<<<<< HEAD
#char_allegiances <- tibble(name = names_vector1, allegiances = allegiances_count)
=======
>>>>>>> c5a67b865174017dd818a26e97b53c1814b97bf7
char_allegiances <- tibble(name = names1, allegiances = allegiances_count)

# Find the character with the most allegiances
max_allegiances_char <- char_allegiances %>% filter(allegiances == max(allegiances))
print(max_allegiances_char)

# Find the character with the least allegiances
min_allegiances_char <- char_allegiances %>% filter(allegiances == min(allegiances))
print(min_allegiances_char)
```

**3c.** Run the code below in your console. What does it do?

```{r, eval=FALSE}
1:5 %in% 3:6
```

Using the logic you can infer about the `%in%` operator (you can also
read its help file), craft a single line of code to compute a Boolean
vector of length 6 that checks whether the first Game of Thrones
character, stored in `got_chars[[1]]`, has appeared in each of the 6 TV
seasons. Print the result to the console.

```{r q3c}
print(1:6 %in% got_chars[[1]]$alive)
```

**3d.** Run the two lines of code below in their console. What do they
do?

```{r, eval=FALSE}
rbind(1:5, 6:10, 11:15)
do.call(rbind, list(1:5, 6:10, 11:15))
```

Using the logic you can infer about the `do.call()` function (you can
also read its help file), as well as the logic from the last question,
complete the following task. Using `map()`, a custom-defined function,
as well as some post-processing of its results, produce a matrix that
has dimension 30 x 6, with each column representing a TV season, and
each row a character. The matrix should have a value of `TRUE` in
position (i,j) if character i was in season j, and `FALSE` otherwise.
Print the first 6 rows of the result to the console.

```{r q3d}
library(purrr)

appear_in_season <- function(character) {
    sapply(1:6, function(season) season %in% character$alive)
}

season_matrix <- map(got_chars, appear_in_season)

season_matrix <- do.call(rbind, season_matrix)

print(season_matrix[1:6, ])
```

**Challenge.** Repeat the same task as in the last question, but using
`map_df()` and no post-processing. The result will now be a data frame
(not a matrix). Print the first 6 rows of the result to the console.
Hint: `map_dfr()` will throw an error if it can't infer column names.

```{r q3chal}
library(purrr)
library(dplyr)

appear_in_season_df <- function(character) {
    data.frame(season1 = 1 %in% character$alive,
               season2 = 2 %in% character$alive,
               season3 = 3 %in% character$alive,
               season4 = 4 %in% character$alive,
               season5 = 5 %in% character$alive,
               season6 = 6 %in% character$alive)
}

# Use map_df to apply this function to each character and combine the results into a data frame
season_df <- map_df(got_chars, appear_in_season_df)

print(season_df[1:6, ])
```

# Q4. Cultural studies

**4a.** Using `map_dfr()`, create a data frame of dimension 30 x 5,
whose columns represent, for each Game of Thrones character, their name,
birth date, death date, gender, and culture. Store it as `got_df` and
print the last 3 rows to the console.

```{r q4a}
library(purrr)

extract_info <- function(character) {
    data.frame(
        name = character$name,
        birth_date = character$born,
        death_date = character$died,
        gender = character$gender,
        culture = character$culture
    )
}

# Use map_dfr to apply this function to each character and combine the results into a data frame
got_df <- map_dfr(got_chars, extract_info)

print(got_df[(nrow(got_df)-2):nrow(got_df), ])
```

**4b.** Using `got_df`, show that you can compute whether each character
is alive or not, and compare this to what is stored in `got_chars`,
demonstrating that the two ways of checking whether each character is
alive lead to equal results.

```{r q4b}
library(purrr)
library(dplyr)

# Step 1: Compute alive status from got_df
is_alive_df <- is.na(got_df$died)

# Step 2: Compute alive status from got_chars
is_alive_chars <- map_lgl(got_chars, function(character) is.na(character$died))

# Step 3: Compare the two vectors
comparison <- is_alive_df == is_alive_chars
print(comparison)
```

**4c.** Using `filter()`, print the subset of the rows of `got_df` that
correspond to Ironborn characters. Then print the subset that correspond
to female Northmen.

```{r q4c}
library(dplyr)

# Filter for Ironborn characters
ironborn <- filter(got_df, culture == "Ironborn")
print(ironborn)

# Filter for female Northmen
female_northmen <- filter(got_df, culture == "Northmen" & gender == "Female")
print(female_northmen)
```

**4d.** Create a matrix of dimension (number of cultures) x 2 that
counts how many women and men there are in each culture appearing in
`got_df`. Print the results to the console. Hint: what happens if you
pass `table()` two arguments?

```{r q4d}
# Create a contingency table for gender within each culture
gender_culture_matrix <- table(got_df$culture, got_df$gender)

print(gender_culture_matrix)
```

**4e.** Using `group_by()` and `summarize()` on `got_df`, compute how
many characters in each culture have died. Which culture---aside from
the unknown category represented by ""---has the most deaths?

```{r q4e}
library(dplyr)

# Compute the number of deceased characters in each culture
deaths_by_culture <- got_df %>%
  group_by(culture) %>%
  summarize(deaths = sum(!is.na(death_date), na.rm = TRUE)) %>%
  filter(culture != "")  # Exclude unknown category represented by ""
print(deaths_by_culture)

# Find the culture with the most deaths
most_deaths_culture <- deaths_by_culture %>%
  filter(deaths == max(deaths))

print(most_deaths_culture)
```

# Rio Olympics data set

This is data set from the Rio Olympics data set that we saw in Lab 3. In
the next question, we're going to repeat some calculations from Lab 3
but using `dplyr`.

```{r}
rio = read.csv("http://www.stat.cmu.edu/~ryantibs/statcomp/data/rio.csv")
```

# Q5. Practice with grouping and summarizing

**5a.** Using `group_by()` and `summarize()`, compute how many athletes
competed for each country in the `rio` data frame? Print the results for
the first 10 countries to the console. Building off your here answer,
use an additional call to `filter()` to compute which country had the
most number of athletes and how many that was. Hint: consider using
`n()` from the `dplyr` package for the first part here.

```{r q5a}
<<<<<<< HEAD
# YOUR CODE GOES HERE

# Display column names
cat("Column Names:\n")
print(names(rio))
cat("\n")

# Display the first row of the data frame
cat("First Row:\n")
print(head(rio, n = 1))


=======
# Step 2 & 3: Group by country and summarize to count athletes
athletes_per_country <- rio %>%
  group_by(nationality) %>%
  summarize(Num_Athletes = n())

# Step 4: Print results for the first 10 countries
print(athletes_per_country[1:10, ])

# Step 5: Identify the country with the most athletes
most_athletes <- athletes_per_country %>%
  filter(Num_Athletes == max(Num_Athletes))

# Print the country with the most athletes and the count
print(most_athletes)
>>>>>>> c5a67b865174017dd818a26e97b53c1814b97bf7
```

**5b.** Using `group_by()`, `summarize()`, and `filter()`, compute which
country had the largest number of total medals and many that was.

```{r q5b}
<<<<<<< HEAD
# YOUR CODE GOES HERE

top_medal_county <- rio %>%
  group_by(nationality) %>%
  summarize(total_medals = sum(gold + silver + bronze)) %>%
  #arrange(desc(total_medals)) %>%
  arrange(total_medals) %>%
  head(5)

# Now print the result
print(top_medal_county)
=======
# Group by country and summarize to count total medals
total_medals_per_country <- rio %>%
  group_by(nationality) %>%
  summarize(Total_Medals = sum(gold, silver, bronze, na.rm = TRUE)) %>%
  arrange(desc(Total_Medals))

# Print results for the country with the most medals
most_medals_country <- total_medals_per_country %>%
  filter(Total_Medals == max(Total_Medals))

# Print the country with the most medals
print(most_medals_country)
>>>>>>> c5a67b865174017dd818a26e97b53c1814b97bf7

```

**5c.** Using `group_by()`, `summarize()`, and `filter()`, compute which
country---among those with zero total medals---had the most number of
athletes. Hint: you will need to modify your `summarize()` command to
compute the number of athletes; and you might need two calls to
`filter()`.

```{r q5c}
<<<<<<< HEAD
# YOUR CODE GOES HERE

result <- rio %>%
  group_by(nationality) %>%
  summarize(total_medals = sum(gold + silver + bronze), num_athletes = n(), .groups = "drop") %>%
  filter(total_medals == 0) %>%
  arrange(desc(num_athletes)) %>%
  head(5)



# Print the result
print(result)
=======
country_medals_athletes <- rio %>%
  group_by(nationality) %>%
  summarize(
    Total_Medals = sum(gold, silver, bronze, na.rm = TRUE),
    Num_Athletes = n()
  ) %>%
  arrange(desc(Num_Athletes))

# First filter to get countries with zero total medals
countries_with_zero_medals <- country_medals_athletes %>%
  filter(Total_Medals == 0)

# Second filter to find the one with the most athletes
country_with_most_athletes_zero_medals <- countries_with_zero_medals %>%
  filter(Num_Athletes == max(Num_Athletes))

# Print the country with the most athletes among those with zero medals
print(country_with_most_athletes_zero_medals)
>>>>>>> c5a67b865174017dd818a26e97b53c1814b97bf7
```

**5d.** Using---yes, you guessed it---`group_by()`, `summarize()`, and
`filter()`, compute the average weight of athletes in each sport,
separately for men and women, and report the two sport with the highest
average weights (one for each of men and women). Hint: `group_by()` can
accept more than one grouping variable. Also, consider using
`na.rm=TRUE` as an additional argument to certain arithmetic summary
functions so that they will not be thrown off by `NA` or `NaN` values.

```{r q5d}
<<<<<<< HEAD
# YOUR CODE GOES HERE

rio %>%
  filter(sex == "male") %>%
  group_by(sport) %>%
  summarize(avg_weight = mean(weight, na.rm = TRUE)) %>%
  arrange(desc(avg_weight)) %>%
  head(5)
  
  
rio %>%
  filter(sex == "female") %>%
  group_by(sport) %>%
  summarize(avg_weight = mean(weight, na.rm = TRUE)) %>%
  arrange(desc(avg_weight)) %>%
  head(5)

=======
# Group by sport and sex, then calculate average weight
average_weight_by_sport_sex <- rio %>%
  group_by(sport, sex) %>%
  summarise(Average_Weight = mean(weight, na.rm = TRUE), .groups = 'drop') %>%
  arrange(desc(Average_Weight))

print(average_weight_by_sport_sex)

# Find the sport with the highest average weight for men
highest_avg_weight_sport_men <- average_weight_by_sport_sex %>%
  filter(sex == "male") %>%
  top_n(1, Average_Weight)

print(highest_avg_weight_sport_men)

# Find the sport with the highest average weight for women
highest_avg_weight_sport_women <- average_weight_by_sport_sex %>%
  filter(sex == "female") %>%
  top_n(1, Average_Weight)

print(highest_avg_weight_sport_women)

# Combine the results
highest_avg_weight_sports <- bind_rows(highest_avg_weight_sport_men, highest_avg_weight_sport_women)

# Print the sports with the highest average weight for men and women
print(highest_avg_weight_sports)
>>>>>>> c5a67b865174017dd818a26e97b53c1814b97bf7
```

# Fastest 100m sprint times

Below, we read two data sets of the 1000 fastest times ever recorded for
the 100m sprint, in men's and women's track. We scraped this data from
<http://www.alltime-athletics.com/m_100ok.htm> and
<http://www.alltime-athletics.com/w_100ok.htm>, in early September 2021.
(Interestingly, the 2nd, 3rd, 4th, 7th, and 8th fastest women's times
were all set at the most recent Tokyo Olympics, or after! Meanwhile, the
top 10 men's times are all from about a decade ago.)

# Prostate cancer data set

Below we read in the prostate cancer data set, as visited in previous
labs.

```{r}
pros.df = 
  read.table("http://www.stat.cmu.edu/~ryantibs/statcomp/data/pros.dat")
```

# Q6. Practice with `dplyr` verbs

In the following, use pipes and `dplyr` verbs to answer questions on
`pros.df`.

**6a.** Among the men whose `lcp` value is equal to the minimum value
(across the entire data set), report the range (min and max) of `lpsa`.

```{r q6a}
# YOUR CODE GOES HERE

# Calculate the desired results using dplyr
result <- pros.df %>%
  summarise(min_lpsa = min(lpsa, na.rm = TRUE),
            max_lpsa = max(lpsa, na.rm = TRUE))

# Print the result
print(result)

```

**6b.** Order the rows by decreasing `age`, then display the rows from
men who are older than 70 and without SVI.

```{r q6b}
# YOUR CODE GOES HERE

# Sort the rows by decreasing age
sorted_pros <- pros.df[order(-pros.df$age, pros.df$svi), ]

# Display the rows for men aged over 70 without SVI
subset_rows <- sorted_pros[(sorted_pros$age > 70 & sorted_pros$svi == 0), ]
head(subset_rows)

```

**6c.** Order the rows by decreasing `age`, then decreasing `lpsa`
score, and display the rows from men who are older than 70 and without
SVI, but only the `age`, `lpsa`, `lcavol`, and `lweight` columns. Hint:
`arrange()` can take two arguments, and the order you pass in them
specifies the priority.

```{r q6c}
# YOUR CODE GOES HERE

# Arrange the rows by decreasing age and decreasing lpsa score
ordered_pros <- pros.df %>%
  arrange(desc(age), desc(lpsa))

# Select rows for men aged over 70 without SVI, showing the selected columns
selected_columns <- ordered_pros %>%
  filter(age > 70 & svi == 0) %>%
  select(age, lpsa, lcavol, lweight)

# Show the resulting table
print(selected_columns)

```

**6d.** We're going to resolve Q2c from Lab 3 using the tidyverse. Using
`purrr` and `dplyr`, perform t-tests for each variable in the data set,
between SVI and non-SVI groups. To be precise, you will perform a t-test
for each column excluding the SVI variable itself, by running the
function `t.test.by.ind()` below (which is just as in Q2c in Lab 3).
Print the returned t-test objects out to the console.

```{r}
t.test.by.ind = function(x, ind) {
  
  stopifnot(all(ind %in% c(0, 1)))
  
  return(t.test(x[ind == 0], x[ind == 1]))
} 
```

```{r q6d}

for (col in names(pros.df)[-which(names(pros.df)=="svi")]) {
  cat("Testing:", col, "\n")
  result <- t.test.by.ind(pros.df[[col]], pros.df$svi)
  print(result)
  cat("\n\n")
}


```

**6e.** Extend your code from the last part (append just one more line
of code, glued together by a pipe) to extract the p-values from each of
the returned t-test objects, and print them out to the console.

```{r q6e}
# YOUR CODE GOES HERE

p_values <- unlist(lapply(
  for (col in names(pros.df)[-which(names(pros.df)=="svi")]) {
      cat("Testing:", col, "\n")
      result <- t.test.by.ind(pros.df[[col]], pros.df$svi) %>%
        broom::tidy() %>%
        dplyr::select(p.value) %>%
        pluck(1)
      
      print(result)
      cat("\n\n")
      result
  },
  FUN=identity
))




```

```{r}
sprint.m.df = read.table(
  file="http://www.stat.cmu.edu/~ryantibs/statcomp/data/sprint.m.txt", 
  sep="\t", quote="", header=TRUE)
sprint.w.df = read.table(
  file="http://www.stat.cmu.edu/~ryantibs/statcomp/data/sprint.w.txt", 
  sep="\t", quote="", header=TRUE)

print(sprint.m.df)
print(sprint.w.df)
```

# Q7 - More practice with data frame computations

**7a.** Confirm that both `sprint.m.df` and `sprint.w.df` are data
frames. Delete the `Rank` column from each data frame, then display the
first and last 3 rows of each.

```{r q7a}
# YOUR CODE GOES HERE

# Check if they are data frames
class(sprint.m.df); class(sprint.w.df)

# Remove Rank column
sprint.m.df <- sprint.m.df[,-1]; sprint.w.df <- sprint.w.df[,-1]

# Display first 3 rows of each data frame
head(sprint.m.df, n = 3); head(sprint.w.df, n = 3)

# Display last 3 rows of each data frame
tail(sprint.m.df, n = 3); tail(sprint.w.df, n = 3)

```

**7b.** Recompute the ranks for the men's data set from the `Time`
column and add them back as a `Rank` column to `sprint.m.df`. Do the
same for the women's data set. After adding back the rank columns, print
out the first 10 rows of each data frame, but only the `Time`, `Name`,
`Date`, and `Rank` columns. Hint: consider using `rank()`.

```{r q7b}
# YOUR CODE GOES HERE
# Compute ranks for males
sprint.m.df$Rank_male <- rank(sprint.m.df$Time)

# Compute ranks for female
sprint.w.df$Rank_female <- rank(sprint.w.df$Time)

# Print out the first 10 rows of each data frame with specified columns
head(sprint.m.df[, c("Time", "Name", "Date", "Rank_male")], n = 10)
head(sprint.w.df[, c("Time", "Name", "Date", "Rank_female")], n = 10)

```

**7c.** Using base R functions, compute, for each country, the number of
sprint times from this country that appear in the men's data set. Call
the result `sprint.m.counts`. Do the same for the women's data set, and
call the result `sprint.w.counts`. What are the 5 most represented
countries, for the men, and for the women? (Interesting side note: go
look up the population of Jamaica, compared to that of the US. Pretty
impressive, eh?)

```{r q7c}
# YOUR CODE GOES HERE

# Count for men
sprint.m.counts <- table(sprint.m.df$Country)

# Count for women
sprint.w.counts <- table(sprint.w.df$Country)


# Top 5 countries for men
top_men <- head(sort(sprint.m.counts, decreasing = TRUE), 5)

# Top 5 countries for women
top_women <- head(sort(sprint.w.counts, decreasing = TRUE), 5)

cat("Top 5 countries for men:\n")
print(top_men)
cat("\nTop 5 countries for women:\n")
print(top_women)

```

**7d.** Repeat the same calculations as in last part but using `dplyr`
functions, and print out again the 5 most represented countries for men
and women. (No need to save new variables.) Hint: consider using
`arrange()` from the `dplyr` library.

```{r q7d}
# YOUR CODE GOES HERE

# Count for men using dplyr
sprint.m.counts.dplyr <- sprint.m.df %>%
  group_by(Country) %>%
  tally() %>%
  arrange(desc(n))

# Count for women using dplyr
sprint.w.counts.dplyr <- sprint.w.df %>%
  group_by(Country) %>%
  tally() %>%
  arrange(desc(n))


cat("Top 5 countries for men (using dplyr): \n")
print(head(sprint.m.counts.dplyr, n = 5))
cat("\n")

cat("Top 5 countries for women (using dplyr): \n")
print(head(sprint.w.counts.dplyr, n = 5))


```

**7e.** Are there any countries that are represented by women but not by
men, and if so, what are they? Vice versa, represented by men and not
women? Hint: consider using the `%in%` operator.

```{r q7e}
# YOUR CODE GOES HERE

countries_only_women <- anti_join(sprint.w.df %>% distinct(Country), sprint.m.df %>% distinct(Country), by = "Country") %>% pull(Country)
countries_only_men <- anti_join(sprint.m.df %>% distinct(Country), sprint.w.df %>% distinct(Country), by = "Country") %>% pull(Country)


cat("Countries represented by women but not by men:", paste(countries_only_women, collapse = ", "), "\n")
cat("Countries represented by men but not by women:", paste(countries_only_men, collapse = ", "), "\n")


```

# Q8. More practice with `dplyr` functions

**8a.** Using `dplyr` functions, compute, for each country, the fastest
time among athletes who come from that country. Do this for each of the
men's and women's data sets, and display the first 10 rows of the
result.

```{r q8a}
# Convert Time to numeric and compute the fastest time for men's dataset
fastest_men <- sprint.m.df %>%
  mutate(Time = as.numeric(gsub("\"", "", Time))) %>%  # Remove quotes and convert to numeric
  group_by(Country) %>%
  summarize(Fastest_Time = min(Time, na.rm = TRUE), .groups = 'drop') %>%
  slice_head(n = 10)

# Convert Time to numeric and compute the fastest time for women's dataset
fastest_women <- sprint.w.df %>%
  mutate(Time = as.numeric(gsub("\"", "", Time))) %>%  # Remove quotes and convert to numeric
  group_by(Country) %>%
  summarize(Fastest_Time = min(Time, na.rm = TRUE), .groups = 'drop') %>%
  slice_head(n = 10)

# Print the results
print(fastest_men)
print(fastest_women)
```

**8b.** With the most minor modification to your code possible, do the
same computations as in the last part, but now display the first 10
results ordered by increasing time. Hint: recall `arrange()`.

```{r q8b}
# Convert Time to numeric and compute the fastest time for men's dataset
fastest_men <- sprint.m.df %>%
  mutate(Time = as.numeric(gsub("\"", "", Time))) %>%  # Remove quotes and convert to numeric
  group_by(Country) %>%
  summarize(Fastest_Time = min(Time, na.rm = TRUE), .groups = 'drop') %>%
  arrange(Fastest_Time) %>%
  slice_head(n = 10)

# Convert Time to numeric and compute the fastest time for women's dataset
fastest_women <- sprint.w.df %>%
  mutate(Time = as.numeric(gsub("\"", "", Time))) %>%  # Remove quotes and convert to numeric
  group_by(Country) %>%
  summarize(Fastest_Time = min(Time, na.rm = TRUE), .groups = 'drop') %>%
  arrange(Fastest_Time) %>%
  slice_head(n = 10)

# Print the results
print(fastest_men)
print(fastest_women)
```

**8c.** Rewrite your solution in the last part using base R. Hint:
`tapply()` gives probably the easiest route here. Note: your code here
shouldn't be too much more complicated than your code in the last part.

```{r q8c}
# Convert Time to numeric after removing quotes for men's dataset
sprint.m.df$Time <- as.numeric(gsub("\"", "", sprint.m.df$Time))

# Use tapply to compute the fastest time for each country in the men's dataset
fastest_times_men <- tapply(sprint.m.df$Time, sprint.m.df$Country, min, na.rm = TRUE)

# Convert the result to a data frame, sort it, and get the first 10 entries
fastest_times_men_df <- data.frame(Country = names(fastest_times_men), Fastest_Time = fastest_times_men)
fastest_times_men_df <- fastest_times_men_df[order(fastest_times_men_df$Fastest_Time), ]
head(fastest_times_men_df, 10)

# Convert Time to numeric after removing quotes for women's dataset
sprint.w.df$Time <- as.numeric(gsub("\"", "", sprint.w.df$Time))

# Use tapply to compute the fastest time for each country in the women's dataset
fastest_times_women <- tapply(sprint.w.df$Time, sprint.w.df$Country, min, na.rm = TRUE)

# Convert the result to a data frame, sort it, and get the first 10 entries
fastest_times_women_df <- data.frame(Country = names(fastest_times_women), Fastest_Time = fastest_times_women)
fastest_times_women_df <- fastest_times_women_df[order(fastest_times_women_df$Fastest_Time), ]
head(fastest_times_women_df, 10)
```

**8d.** Using `dplyr` functions, compute, for each country, the
quadruple: name, city, country, and time, corresponding to the athlete
with the fastest time among athletes from that country. Do this for each
of the men's and women's data sets, and display the first 10 rows of the
result, ordered by increasing time. If there are ties, then show all the
results that correspond to the fastest time. Hint: consider using
`select()` from the `dplyr` library.

```{r q8d}
# For the men's dataset
fastest_men <- sprint.m.df %>%
  mutate(Time = as.numeric(gsub("\"", "", Time))) %>%
  arrange(Country, Time) %>%
  group_by(Country) %>%
  filter(Time == min(Time, na.rm = TRUE)) %>%
  select(Name, City, Country, Time) %>%
  ungroup() %>%
  arrange(Time) %>%
  slice_head(n = 10)

# For the women's dataset
fastest_women <- sprint.w.df %>%
  mutate(Time = as.numeric(gsub("\"", "", Time))) %>%
  arrange(Country, Time) %>%
  group_by(Country) %>%
  filter(Time == min(Time, na.rm = TRUE)) %>%
  select(Name, City, Country, Time) %>%
  ungroup() %>%
  arrange(Time) %>%
  slice_head(n = 10)

# Print the results
print(fastest_men)
print(fastest_women)

```

**8e.** Rewrite your solution in the last part using base R. Hint: there
are various routes to go; one strategy is to use `split()`, followed by
`lapply()` with a custom function call, and then `rbind()` to get things
in a data frame form. Note: your code here will probably be more
complicated, or at least less intuitive, than your code in the last
part.

```{r q8e}
# Assuming sprint.m.df and sprint.w.df have been loaded into the R environment

# Define a custom function to find the fastest athlete per country
fastest_athlete <- function(df) {
  df$Time <- as.numeric(gsub("\"", "", df$Time))
  min_time <- min(df$Time, na.rm = TRUE)
  fastest <- df[df$Time == min_time, ]
  return(fastest)
}

# Apply the function to each subset of the data split by country
fastest_men <- do.call(rbind, lapply(split(sprint.m.df, sprint.m.df$Country), fastest_athlete))
fastest_women <- do.call(rbind, lapply(split(sprint.w.df, sprint.w.df$Country), fastest_athlete))

# Order the results by increasing time and select the first 10 rows
fastest_men_ordered <- fastest_men[order(fastest_men$Time), ][1:10, ]
fastest_women_ordered <- fastest_women[order(fastest_women$Time), ][1:10, ]

# Print the results
print(fastest_men_ordered)
print(fastest_women_ordered)
```

**8f.** Order the rows by increasing `Wind` value, and then display only
the women who ran at most 10.7 seconds.

```{r q8f}
# Convert Time to numeric after removing quotes for the women's dataset
sprint.w.df$Time <- as.numeric(gsub("\"", "", sprint.w.df$Time))

# Order by Wind and then filter for times at most 10.7 seconds
fastest_women_under_10_7 <- sprint.w.df[sprint.w.df$Time <= 10.7, ]
fastest_women_under_10_7_ordered <- fastest_women_under_10_7[order(fastest_women_under_10_7$Wind), ]

# Display the result
print(fastest_women_under_10_7_ordered)
```

**8g.** Order the rows by terms of increasing `Time`, then increasing
`Wind`, and again display only the women who ran at most 10.7 seconds,
but only the `Time`, `Wind`, `Name`, and `Date` columns.

```{r q8g}
# Assuming sprint.w.df has been loaded into the R environment

# Convert Time to numeric after removing quotes
sprint.w.df$Time <- as.numeric(gsub("\"", "", sprint.w.df$Time))

# Order by Time, then Wind, filter for times at most 10.7 seconds, and select columns
fastest_women <- sprint.w.df[sprint.w.df$Time <= 10.7, ]
fastest_women_ordered <- fastest_women[order(fastest_women$Time, fastest_women$Wind), ]
selected_columns <- fastest_women_ordered[c("Time", "Wind", "Name", "Date")]

# Display the result
print(selected_columns)
```

**8h.** Plot the `Time` versus `Wind` columns, but only using data where
`Wind` values that are nonpositive. Hint: note that for a data frame,
`df` with columns `colX` and `colY`, you can use
`plot(colY ~ colX, data=df)`, to plot `df$colY` (y-axis) versus
`df$colX` (x-axis).

```{r q8h}
# Convert Time to numeric after removing quotes
sprint.w.df$Time <- as.numeric(gsub("\"", "", sprint.w.df$Time))

# Subset the data where Wind values are nonpositive
nonpositive_wind_data <- sprint.w.df[sprint.w.df$Wind <= 0, ]

# Plot Time versus Wind
plot(Time ~ Wind, data=nonpositive_wind_data,
     main="Time versus Wind for Nonpositive Wind Values",
     xlab="Wind (m/s)", ylab="Time (seconds)",
     pch=19, col="blue")
```

**8i.** Extend your code from the last part (append just two more lines
of code, glued together by a pipe) to plot the single fastest `Time` per
`Wind` value. (That is, your plot should be as in the last part, but
among points that share the same x value, only the point with the lowest
y value should be drawn.)

```{r q8i}
# Assuming sprint.w.df has been loaded into the R environment

# Convert Time to numeric after removing quotes
sprint.w.df$Time <- as.numeric(gsub("\"", "", sprint.w.df$Time))

# Subset the data where Wind values are nonpositive
nonpositive_wind_data <- sprint.w.df[sprint.w.df$Wind <= 0, ]

# Find the fastest Time for each Wind value
fastest_time_per_wind <- aggregate(Time ~ Wind, data = nonpositive_wind_data, min)

# Plot the fastest Time versus Wind
plot(Time ~ Wind, data = fastest_time_per_wind,
     main = "Fastest Time versus Wind for Nonpositive Wind Values",
     xlab = "Wind (m/s)", ylab = "Fastest Time (seconds)",
     pch = 19, col = "blue")

```

# Q9. Practice pivoting wider and longer

In the following, use pipes and `dplyr` and `tidyr` verbs to answer
questions on `sprint.m.df`. In some parts, it might make more sense to
use direct indexing, and that's perfectly fine.

**9a.** Confirm that the `Time` column is stored as character data type.
Why do you think this is? Convert the `Time` column to numeric. Hint:
after converting to numeric, there will be `NA` values; look at the
position of one such `NA` value and revisit the original `Time` column
to see why it was stored as character type in the first place.

```{r q9a}
# Confirm the data type of the Time column
class(sprint.m.df$Time)

# Convert the Time column to numeric, this will create NA values for non-convertible entries
sprint.m.df <- sprint.m.df %>%
  mutate(Time_numeric = as.numeric(gsub("\"", "", Time)))

# Find the position of the NA values in the new Time_numeric column
na_positions <- which(is.na(sprint.m.df$Time_numeric))

# Inspect the original Time column for an NA position
sprint.m.df$Time[na_positions]

```

**9b.** Define a reduced data frame `dat.reduced` as follows. For each
athlete, and each city, keep the fastest of all times they recorded in
this city. Then drop all rows with an `NA` value in the `Time` column
Your new data frame `dat.reduced` should have 600 rows and 3 columns
(`Name`, `City`, `Time`). Confirm that it has these dimensions, and
display its first 10 rows. Hint: `drop_na()` in the `tidyr` package
allows you to drop rows based on `NA` values.

```{r q9b}
# Convert Time to numeric (handling the conversion issue)
sprint.m.df <- sprint.m.df %>%
  mutate(Time = as.numeric(gsub("\"", "", Time, fixed = TRUE)))

# Create dat.reduced
dat.reduced <- sprint.m.df %>%
  group_by(Name, City) %>%
  summarize(Time = min(Time, na.rm = TRUE), .groups = 'drop') %>%
  drop_na(Time)

# Confirm dimensions and display the first 10 rows
print(dim(dat.reduced))
print(head(dat.reduced, 10))

```

**9c.** The data frame `dat.reduced` is said to be in "long" format: it
has observations on the rows, and variables (`Name`, `City`, `Time`) on
the columns. Arrange the rows alphebtically by city; convert this data
frame into "wide" format; and then order the rows so that they are
alphabetical by sprinter name. Call the result `dat.wide`. To be clear,
here the first column should be the athlete names, and the remaining
columns should correspond to the cities. Confirm that your data frame
has dimension 141 x 152 Do these dimensions make sense to you?

```{r q9c}
dat.wide <- dat.reduced %>%
  arrange(City) %>%
  pivot_wider(names_from = City, values_from = Time) %>%
  arrange(Name)

# Confirm dimensions
print(dim(dat.wide))
```

**9d.** Not counting the names in the first column, how many non-`NA`
values does `dat.wide` have? How could you have guessed this number
ahead of time, directly from `dat.reduced` (before any pivoting at all)?

```{r q9d}
non_na_count <- sum(!is.na(dat.wide[, -1])) # Exclude the first column (Name)
print(non_na_count)

# How to guess this number from dat.reduced
print(nrow(dat.reduced)) # This should match non_na_count
```

**9e.** From `dat.wide`, look at the row for "Usain Bolt", and determine
the city names that do not have `NA` values. These should be the cities
in which he raced. Determine these cities directly from `dat.reduced`,
and confirm that they match.

```{r q9e}
usain_bolt_cities_wide <- names(dat.wide)[which(!is.na(dat.wide[dat.wide$Name == "Usain Bolt", -1])) + 1]
print(usain_bolt_cities_wide)

# Directly from dat.reduced
usain_bolt_cities_reduced <- dat.reduced %>%
  filter(Name == "Usain Bolt") %>%
  pull(City)
print(usain_bolt_cities_reduced)

# Confirm they match
print(all(usain_bolt_cities_wide == usain_bolt_cities_reduced))
```

**9f.** Convert `dat.wide` back into "long" format, and call the result
`dat.long`. Remove rows that have `NA` values (hint: you can do this by
setting `values_drop_na = TRUE` in the call to the pivoting function),
and order the rows alphabetically by athlete and city name. Once you've
done this, `dat.long` should have matching entries to `dat.reduced`;
confirm that this is the case.

```{r q9f}
dat.long <- dat.wide %>%
  pivot_longer(-Name, names_to = "City", values_to = "Time", values_drop_na = TRUE) %>%
  arrange(Name, City)

# Confirm matching entries with dat.reduced
print(all.equal(dat.long, dat.reduced))
```
